Cassini State Capture

### 2019-08-07 writeups/Su_sep_cross.tex
temp checkpoint

The area enclosed by the separatrix is known exactly in literature
\citep{henrard1987}, but a serviceable approximation for small $\eta$ can be
derived as follows. Call $H_4$ the value of $H$ at CS4, then the separatrix is
defined implicitly by solutions to $H\p{\mu_{sep}(\phi), \phi} = H_4$. This may
be evaluated and we obtain
\begin{equation}
    \mu_{sep}(\phi) \approx \pm \sqrt{2\eta \sin I \p{1 - \cos \phi}}
        + \eta \cos I + \mathcal{O}\p{\eta^{3/2}}.\label{eq:mu_sep}
\end{equation}
The enclosed area between the two solutions can be obtained via explicit
integration. It can be expressed as a fraction of the total phase space area:
\begin{equation}
    \frac{A_{sep}}{4\pi} \approx \frac{4\sqrt{\eta \sin I}}{\pi}
        + \mathcal{O}\p{\eta^{3/2}}.\label{eq:a_sep}
\end{equation}

@@

Consider our Hamiltonian described in \autoref{eq:H} as a unperturbed
Hamiltonian $H^{(0)}(\mu, \phi; \eta)$ where $\eta$ is to be varied
adiabatically. As $\eta$ varies, the location of the separatrix also changes. We
follow~\cite{henrard1982} by defining
\begin{equation}
    h(\mu, \phi; \eta) \equiv H^{(0)}(\mu, \phi; \eta) - H_4(\eta).
\end{equation}
The significance of $h$ is that it always vanishes along the separatrix. Note
that $h$ corresponds to $K$ in~\cite{henrard1982}, though in our problem
$H^{(0)} > H_4$ corresponds to the \emph{interior} of the separatrix. The
evolution of $h$ over time then determines whether and when a trajectory
experiences separatrix crossing. Denoting time derivatives with dots, we write
\begin{align}
    \rd{h}{t} &= \rd{H^{(0)}}{t} - \pd{H_4}{\eta}\dot{\eta},\nonumber\\
        &= \s{\pd{H^{(0)}}{\mu}\dot{\mu}
                + \pd{H^{(0)}}{\phi}\dot{\phi}
                + \pd{H^{(0)}}{\eta}\dot{\eta}}
            - \pd{H_4}{\eta}\dot{\eta},\nonumber\\
        &= \s{\pd{H^{(0)}}{\mu}\dot{\mu}
                + \pd{H^{(0)}}{\phi}\dot{\phi}
                + \pd{H^{(0)}}{\eta}\dot{\eta}}
            - \pd{H_4}{\eta}\dot{\eta},\nonumber\\
        &= \s{\dot{\phi}^{(0)}\dot{\mu}^{(1)}
                - \dot{\mu}^{(0)}\dot{\phi}^{(1)}
                + \pd{H^{(0)}}{\eta}\dot{\eta}} -
            \pd{H_4}{\eta}\dot{\eta}.\label{eq:dhdt}
\end{align}
We've notated $\dot{\phi}^{(0)} \equiv \pd{H^{(0)}}{\mu}$ and $\dot{\phi}^{(1)}
= \dot{\phi} - \dot{\phi}^{(0)}$ and similarly for $\mu$. The scenario studied
in~\cite{henrard1982} corresponds to $\dot{\mu}^{(1)} = \dot{\phi}^{(1)} = 0$
while that in~\cite{g_and_h} sets $\dot{\eta} = 0$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/3zones.png}
    \caption{Nomenclature of the two legs of the separatrix $\mathcal{C}_{\pm}$
    and three zones of the domain. For both $\mathcal{C}_{\pm}$, we will take
    the positive direction to be anti-clockwise.}\label{fig:1_3zones}
\end{figure}
For concreteness, we consider a specific example first. We assign labels to our
phase space as shown in \autoref{fig:1_3zones}. Consider a trajectory that
begins in zone III, which is circulating with $\dot{\phi} > 0$ and has $h < 0$,
and consider if $\rd{h}{t} > 0$ in all of zone III, such that the trajectory
eventually experiences separatrix crossing at $\mathcal{C}_-$. The following
sequence of events unfolds:
\begin{enumerate}
    \item[\textbf{(i)}] At the beginning ($\phi = 0$) of its separatrix-crossing
        orbit, $h_i \equiv h(\mu, \phi = 0; \eta) \in (\Delta_-, 0)$. Note
        that $h_i < 0$ to still be in zone III at the start of the orbit, while
        $h_i > \Delta_-$ to be separatrix crossing during the final orbit.

@@

        \begin{equation}
            P_c \equiv \frac{\Delta_+ + \Delta_-}{\Delta_-}.
                \label{eq:p_c_gen}
        \end{equation}
\end{itemize}
With the exception of \autoref{eq:dhdt}, the above outlines the argument given
in~\cite{henrard1982}.

Similar derivations can be given for trajectories originating in zone I.
TODO say something about zone-II originating systems.

@@

Visually, this is depicted in \autoref{fig:0_6manifolds}. Note that:
trajectories starting in zone III below $W_s^{(1)}$ fail to separatrix cross;
those between $W_s^{(1)}$ and $W_s^{(0)}$ continue between $W_u^{(1)}$ and
$W_s^{(0)}$, eventually escaping; those between $W_s^{(0)}$ and $W_u^{(0)}$ are
securely captured, and no trajectories can begin above $W_u^{(0)}$ at $\phi =
0$.

The saddle connection splitting is easiest to measure where $\vec{\nabla}H^{(0)}
\propto \hat{\phi}$, where $\dot{\mu} = 0$, and \autoref{eq:melnikov} reduces to
just $\Delta \mu \pd{H^{(0)}}{\mu} = \Delta H^{(0)}$. Then the capture
probability must be given by ratio of $\Delta \mu$ separating $W_s^{(0)},
W_u^{(0)}$ to $\Delta \mu$ separating $W_s^{(1)}, W_u^{(0)}$. These $\Delta \mu$
values are those given in the title, and the resultant $P_c$ capture probability
is also quoted.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/6manifolds0_20.png}
    \caption{Sample plot of saddle connection breaking in the presence of weakly
    non-Hamiltonian dissipation. Equations are used from \autoref{s:p1}. Stable
    and unstable manifolds are indexed such that superscript $(0)$/$(1)$
    belongs to the saddle point at left/right. $\Delta \mu$ represent the
    manifold separations, resulting in capture probability $P_c$
    prediction per \autoref{eq:p_c_dmu}. Numerical simulations predict $P_c
    \approx 0.252$.}\label{fig:0_6manifolds}
\end{figure}

Finally, the connection between \autoref{ss:henrard} and \autoref{ss:meln} is
evident: the ratio of the $\Delta \mu$ values is simply
\begin{align}
    P_c &= \frac{\Delta \mu(W_s^{(0)}, W_u^{(0)})}{
            \Delta \mu\p{W_s^{(1)}, W_u^{(0)}}},\label{eq:p_c_dmu},\\
        &= \frac{\Delta H^{(0)}_- + \Delta H^{(0)}_+}{\Delta H^{(0)}_-}.
\end{align}
This is exactly \autoref{eq:p_c_gen} when $\dot{\eta} = 0$.

\section{Problem 1: Weakly Dissipitive Hamiltonian}\label{s:p1}

We first consider a toy separatrix crossing problem in the $\dot{\eta} = 0$
limit, the case discussed in \autoref{ss:meln}. Consider adding to the
Hamiltonian Cassini State problem an additional aligning term
$\p{\rd{\theta}{t}}^{(1)} = -\frac{1}{t_s}\sin\theta$ that favors alignment
$\theta = 0$ on synchronization timescale $t_s$. This translates to equation of
motion
\begin{equation}
    \rd{\hat{s}}{\tau} = \p{\hat{s} \cdot \hat{l}}\p{\hat{s} \times \hat{l}}
        - \eta\p{\hat{s} \times \hat{l}_p}
        - \epsilon \sin \theta \hat{\theta}.
\end{equation}
We have divided \autoref{eq:dsdt_base} by $\alpha$ and defined $\tau \equiv
\alpha t, \epsilon \equiv \frac{1}{\alpha t_s}$. The equations of motion for
canonical variables $\mu \equiv \cos\theta, \phi$ are
\begin{subequations}\label{se:p1_eom}
    \begin{align}
        \mu' &= -\cos\theta
            + \eta\p{\cos I + \sin I \frac{\mu}{\sqrt{1 - \mu^2}} \cos \phi},\\
        \mu' &= -\eta \sin I \sin \theta \sin \phi
            \s{+ \epsilon\p{1 - \mu^2}}.
    \end{align}
\end{subequations}
Primes denote $\rd{}{\tau}$. The bracketed term in $\mu'$ is the perturbation
component.

With a dissipitive term that favors alignment, we expect that separatrix
encounters can only occur from zone II and zone III\@. We will focus on the
outcomes of trajectories origiating in zone III for now. We now aim to compute
\autoref{eq:p_c_gen}, for which we need:
\begin{subequations}
    \begin{align}
        \Delta_- &= \int\limits_{\mathcal{C}_-}
                \phi'\epsilon\p{1 - \mu^2}\;\mathrm{d}t,\nonumber\\
            &= \int\limits_{\mathcal{C}_s}
                \p{1 - \mu^2}\;\mathrm{d}\phi,\\
        \Delta_- + \Delta_+ &= \int\limits_{\mathcal{C}_- + \mathcal{C}_+}
            \epsilon\p{1 - \mu^2}\;\mathrm{d}\phi.
    \end{align}
\end{subequations}
Note that along $\mathcal{C}_{\pm}$ that $\mathrm{d}\phi$ has sign $\mp$
respectively, thanks to our sign convention. We proceed to integrate:
\begin{itemize}
    \item $\Delta_-$: Recalling the parameterization for the separatrix
        $\mu_{sep}(\phi)$ given by \autoref{eq:mu_sep}, we can evaluate
        explicitly over $\mathcal{C}_-$. This gives result
        \begin{equation}
            \Delta_- = 2\pi\p{1 - 2\eta \sin I - \eta^2 \cos^2 I}
                + 16\cos I \eta^{3/2}\sqrt{\sin I}.
        \end{equation}

    \item $\Delta_- + \Delta_+$: Note $\mathrm{d}\phi$ changes signs between
        $\mathcal{C}_-$ and $\mathcal{C}_+$. Thus, only terms in $\Delta_-$ that
        also change signs on $\mathcal{C}_+$ contribute to the integral. This
        allows us to evaluate
        \begin{align}
            \Delta_- + \Delta_+ &= 32\epsilon \eta^{3/2} \cos I \sqrt{\sin I}.
        \end{align}
\end{itemize}
From these two integrals, we obtain
\begin{equation}
    P_c = \frac{16\eta^{3/2}\cos I \sqrt{\sin I}}{\pi
        \p{1 - 2\eta \sin I - \eta^2 \cos^2 I}
        + 8\eta^{3/2}\cos I \sqrt{\sin I}}.\label{eq:p1pc}
\end{equation}

@@

A further interesting feature in \autoref{fig:0_3stats} is a small portion of
points in zone II near CS4  at $\phi = 2\pi$ that escape. These points accrue
$\Delta_+$ on their first orbit and escape without librating. Visually, this
zone is obvious in \autoref{fig:0_6manifolds} as the separation between
$W_s^{(1)} \cup W_u^{(1)}$ and $W_s^{(0)} \cup W_u^{(0)}$ near $\phi = 2\pi$.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_1.png}
    \caption{Plot of the CSs that initial conditions converge to. Each subplot
    has a set of initial conditions (black dots) and the CS (red dot) that each
    of these initial conditions converge to. Initial conditions are chosen
    uniformly in $(\mu, \phi)$ coordinates. The parameters for the simulation
    are in the title.}\label{fig:0_3stats}
\end{figure}

>>>>

The area enclosed by the separatrix is known exactly in literature
\citep{henrard1987}, but a serviceable approximation for small $\eta$ can be
derived as follows. Call $H_{sep}$ the value of $H$ along the separatrix (also
the value of $H$ at CS4), then the separatrix is defined implicitly by solutions
to $H\p{\mu_{sep}(\phi), \phi} = H_{sep}$. This may be evaluated and we obtain
\begin{equation}
    \mu_{sep}(\phi) \approx \pm \sqrt{2\eta \sin I \p{1 - \cos \phi}}
        + \eta \cos I + \mathcal{O}\p{\eta^{3/2}}.\label{eq:mu_sep}
\end{equation}
The enclosed area between the two solutions can be obtained via explicit
integration. It can be expressed as a fraction of the total phase space area:
\begin{equation}
    \frac{A_{sep}}{4\pi} \approx \frac{4\sqrt{\eta \sin I}}{\pi}
        + \mathcal{O}\p{\eta^{3/2}}.\label{eq:a_sep}
\end{equation}

@@

Consider our Hamiltonian described in \autoref{eq:H} as a unperturbed
Hamiltonian $H^{(0)}(\mu, \phi; \eta)$ where $\eta$ is to be varied
adiabatically. As $\eta$ varies, the location of the separatrix also changes. We
follow~\cite{henrard1982} by defining
\begin{equation}
    h(\mu, \phi; \eta) \equiv H^{(0)}(\mu, \phi; \eta) - H_{sep}(\eta).
\end{equation}
The significance of $h$ is that it always vanishes along the separatrix. Note
that $h$ corresponds to $K$ in~\cite{henrard1982}, though in our problem
$H^{(0)} > H_{sep}$ corresponds to the \emph{interior} of the separatrix. The
evolution of $h$ over time then determines whether and when a trajectory
experiences separatrix crossing. Denoting time derivatives with dots, we write
\begin{align}
    \rd{h}{t} &= \rd{H^{(0)}}{t} - \pd{H_{sep}}{\eta}\dot{\eta},\nonumber\\
        &= \s{\pd{H^{(0)}}{\mu}\dot{\mu}
                + \pd{H^{(0)}}{\phi}\dot{\phi}
                + \pd{H^{(0)}}{\eta}\dot{\eta}}
            - \pd{H_{sep}}{\eta}\dot{\eta},\nonumber\\
        &= \s{\pd{H^{(0)}}{\mu}\dot{\mu}
                + \pd{H^{(0)}}{\phi}\dot{\phi}
                + \pd{H^{(0)}}{\eta}\dot{\eta}}
            - \pd{H_{sep}}{\eta}\dot{\eta},\nonumber\\
        &= \s{\dot{\phi}^{(0)}\dot{\mu}^{(1)}
                - \dot{\mu}^{(0)}\dot{\phi}^{(1)}
                + \pd{H^{(0)}}{\eta}\dot{\eta}} -
            \pd{H_{sep}}{\eta}\dot{\eta}.\label{eq:dhdt}
\end{align}
We've notated $\dot{\phi}^{(0)} \equiv \pd{H^{(0)}}{\mu}$ and $\dot{\phi}^{(1)}
= \dot{\phi} - \dot{\phi}^{(0)}$ and similarly for $\mu$. The scenario studied
in~\cite{henrard1982} corresponds to $\dot{\mu}^{(1)} = \dot{\phi}^{(1)} = 0$
while that in~\cite{g_and_h} sets $\dot{\eta} = 0$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/3zones.png}
    \caption{Nomenclature of the two legs of the separatrix $\mathcal{C}_{\pm}$
    and three zones of the domain. For both $\mathcal{C}_{\pm}$, we will take
    the positive direction to be anti-clockwise.}\label{fig:1_3zones}
\end{figure}

\subsection{Zone I/III Initial Conditions}\label{ss:z_out_ic}

For concreteness, we consider a specific example first. We assign labels to our
phase space as shown in \autoref{fig:1_3zones}. Consider a trajectory that
begins in zone III, which is circulating with $\dot{\phi} > 0$ and has $h < 0$,
and consider if $\rd{h}{t} > 0$ in all of zone III, such that the trajectory
eventually experiences separatrix crossing at $\mathcal{C}_-$. The following
sequence of events unfolds:
\begin{enumerate}
    \item[\textbf{(i)}] At the beginning ($\phi = 0$) of its separatrix-crossing
        orbit, $h_i \equiv h(\mu, \phi = 0; \eta) \in (\Delta_-, 0)$. Note
        that $h_i < 0$ to still be in zone III at the start of the orbit, while
        $h_i > \Delta_-$ to be separatrix crossing during the final orbit.

@@

        \begin{equation}
            P_c \equiv \frac{\Delta_+ + \Delta_-}{\Delta_-}.
                \label{eq:p_c_gen}
        \end{equation}
\end{itemize}
With the exception of \autoref{eq:dhdt}, the above outlines the argument given
in~\cite{henrard1982}. Similar conclusions can be drawn for trajectories
originating in zone I.

\subsection{Zone II Initial Conditions}\label{ss:z2_ic}

The fates of trajectories originating in zone II, which librate about the
CS2, also depend on $\Delta_{\pm}$. This implies $h_i > 0$. We identify the
following possibilities:
\begin{itemize}
    \item $\Delta_{+} + \Delta_- > 0$: trajectories near the separatrix are
        driven back into zone II over the course of a full libration, therefore
        any trajectory originating in zone II is bound to zone II\@. Note that
        this alone need not imply convergence to CS2, e.g.\ for a
        non-dissipative system.

    \item $\Delta_+ + \Delta_- < 0, \Delta_+ > 0, \Delta_- < 0$: trajectories
        near the separatrix are driven back into zone II along $\mathcal{C}_+$
        but out of zone II along $\mathcal{C}_-$, the net effect amounting to
        expulsion. Since expulsion cannot occur along $\mathcal{C}_+$, the only
        permitted transition is II $\to$ III\@. Vice versa of course if the signs
        of $\Delta_{\pm}$ are swapped while still satisfying $\Delta_+ +
        \Delta_- < 0$.

    \item $\Delta_\pm < 0$: trajectories near the separatrix tend to be
        ejected along both $\mathcal{C}_{\pm}$. Which zone the trajectory enters
        is then probabilistic: it enters zone I with probability
        $\frac{\Delta_+}{\Delta_+ + \Delta_-}$ and zone III with probability
        $\frac{\Delta_-}{\Delta_+ + \Delta_-}$.

        It bears noting that in the non-dissipative limit, this result is
        equivalent to \citep{henrard1987}, which computes these probabilities
        using change in enclosed phase space area.
\end{itemize}

@@

Visually, this is depicted in \autoref{fig:0_6manifolds}. Note that:
trajectories starting in zone III below $W_s^{R}$ fail to separatrix cross;
those between $W_s^{R}$ and $W_s^{L}$ continue between $W_u^{R}$ and $W_s^{L}$,
eventually escaping; those between $W_s^{L}$ and $W_u^{L}$ are securely
captured, and no trajectories can begin above $W_u^{L}$ at $\phi = 0$.

The saddle connection splitting is easiest to measure where $\vec{\nabla}H^{(0)}
\propto \hat{\phi}$, where $\dot{\mu} = 0$, and \autoref{eq:melnikov} reduces to
just $\Delta \mu \pd{H^{(0)}}{\mu} = \Delta H^{(0)}$. Then the capture
probability must be given by ratio of $\Delta \mu$ separating $W_s^{L},
W_u^{L}$ to $\Delta \mu$ separating $W_s^{R}, W_u^{L}$. These $\Delta \mu$
values are those given in the title, and the resultant $P_c$ capture probability
is also quoted.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/6manifolds0_20.png}
    \caption{Sample plot of saddle connection breaking in the presence of weakly
    non-Hamiltonian dissipation. Equations are used from \autoref{s:p2}. Stable
    and unstable manifolds are indexed such that superscript $L/R$ belongs to
    the saddle point at left/right. $\Delta \mu$ represent the manifold
    separations, resulting in capture probability $P_c$ prediction per
    \autoref{eq:p_c_dmu}. Numerical simulations predict $P_c \approx
    0.252$.}\label{fig:0_6manifolds}
\end{figure}

Finally, the connection between \autoref{ss:henrard} and \autoref{ss:meln} is
evident: the ratio of the $\Delta \mu$ values is simply
\begin{align}
    P_c &= \frac{\Delta \mu(W_s^{L}, W_u^{L})}{
            \Delta \mu\p{W_s^{R}, W_u^{L}}},\label{eq:p_c_dmu},\\
        &= \frac{\Delta H^{(0)}_- + \Delta H^{(0)}_+}{\Delta H^{(0)}_-}.
\end{align}
This is exactly \autoref{eq:p_c_gen} when $\dot{\eta} = 0$.

\section{Problem 1: Adiabatic Capture}\label{s:p1}

Here we will adiabatically vary $\eta$. This problem is extensively covered in
existing work e.g.\ \citep{henrard1987}.

\subsection{Spindown Model, $\dot{\eta} > 0$ (few sims)}\label{ss:p1_spindown}

We may first consider the unperturbed Hamiltonian
system with $\rd{\eta}{t} = \frac{\eta}{t_\eta}$ growth with characteristic
timescale $t_\eta$. Again rescaling time $\tau \equiv \alpha t$ and defining
$\epsilon \equiv \frac{1}{\alpha t_\eta}$, we obtain
\begin{equation}
    \rd{\eta}{\tau} = \epsilon \eta,\label{eq:p1_eom}
\end{equation}
where $\epsilon > 0$. Such a model is motivated e.g.\ when a planet's
spin $\hat{s}$ orbits a host star with angular momentum $\hat{l}$ while the star
spins down; $\dot{\alpha} < 0$ here, and $\eta' > 0$.

We then seek $\Delta_{\pm}$, determined by
\begin{align}
    \Delta_{\pm} &= \int\limits_{\mathcal{C}_{\pm}}
            \pd{h}{\eta}\eta'\;\mathrm{d}\tau,\nonumber\\
        &= \int\limits_{\mathcal{C}_{\pm}}
            \frac{\pd{h}{\eta}}{\phi'}\eta'\;\mathrm{d}\phi.
\end{align}
We must proceed carefully since $\phi' \to 0$ at the endpoints of
$\mathcal{C}_{\pm}$; careful work shows these zeros cancel against zeros of
$\pd{h}{\eta}$, and
\begin{align}
    \Delta_{\pm} &= \int\limits_{\mathcal{C}_{\pm}} \eta'
        \p{\cos I \mp \sin I \sqrt{\frac{1 - \cos \phi}{2\eta\sin I}}}
            \;\mathrm{d}\phi,\label{eq:p1_delta_eq}\\
        &= \mp 2\pi \epsilon \eta \cos I
            + 4\epsilon\sqrt{\eta \sin I}.\label{eq:p1_delta}
\end{align}
Call $\eta_\star$ the value of $\eta$ at which a trajectory encounters the
separatrix. There are then three possibilities:
\begin{itemize}
    \item For $\eta_\star < \frac{4\sin I}{\pi^2 \cos^2 I}$, we obtain
        $\Delta_{\pm} > 0$. In this regime, all separatrix encounters
        originating in zones I/III result in secure capture (see
        \autoref{ss:z_out_ic}), while all trajectories originating in zone II do
        not experience separatrix encounter. For $I = 20^\circ$, this condition
        equates to $\eta_\star \lesssim 0.16$. This condition is not to be taken
        too exactly, as all our calculations are performed in the small $\eta$
        limit. Select simulations in $\eta_\star \lesssim 0.1$ are in agreement
        with the secure capture prediction. An example of such a simulation is
        presented in \autoref{fig:2_1testo4}.
        \begin{figure}[t]
            \centering
            \includegraphics[width=\columnwidth]{../initial/2_toy2/1testo4.png}
            \caption{Simulation with initial condition in zone I subject to
            $\eta' = \epsilon \eta$, where $\epsilon = 10^{-4}$. The top panel
            shows the trajectory (red) starting with initial condition $\mu_0 =
            0.140, \phi_0 = 0$ (green dot) and separatrix at separatrix crossing
            (dashed black) in the labelled coordinates. The end of the
            simulation is labelled in the blue cyan dot. The bottom panel shows
            the evolution of the enclosed area by the trajectory (red dots),
            area of the separatrix (blue) and $\eta$ (dotted green) over
            time.}\label{fig:2_1testo4}
        \end{figure}

    \item If $\eta_\star > \frac{4\sin I}{\pi^2 \cos^2 I}, \eta_\star < \eta_c$,
        then the separatrix still exists, and $\Delta_+ < 0, \Delta_- > 0$,
        $\Delta_+ + \Delta_- > 0$. This implies: zone I + zone II trajectories
        do not experience separatrix encounters, while zone III trajectories are
        captured into zone II with probability
        \begin{equation}
            P_c = \frac{8\sqrt{\eta_\star \sin I}}{2\pi \eta_\star \cos I +
                4\sqrt{\eta_\star \sin I}}.
        \end{equation}

    \item Further evolution from either of the above two cases, as well as the
        case where no separatrix encounter occurs before $\eta > \eta_c$, is
        governed simply by conservation of adiabatic invariant/action variable
        \begin{equation}
            I \equiv \oint \cos \theta \;\mathrm{d}\phi.
        \end{equation}
\end{itemize}

\subsection{Spinup Model, $\dot{\eta} < 0$ (no sims)}\label{ss:p1_spinup}

If instead $\epsilon < 0$, then the separatrix area shrinks over time. For
initial values $\eta > \eta_c$, the separatrix appears over time and the
scenario is analyzed in \citep{anderson2018teeter}.

If $\eta$ decreases past critical value $\frac{4\sin I}{\pi^2 \cos^2 I}$, i.e.\
$\Delta_{\pm} < 0$, then the separatrix appears. Here, trajectories in zones
I/III do not experience separatrix encounter while trajectories in zone II
transition into zones I/III with probabilities given respectively (see
\autoref{ss:z2_ic}):
\begin{equation}
    \begin{split}
        P_{II \to I} &= \frac{2\pi \eta_\star \cos I + 4\sqrt{\eta_\star \sin I}}
            {8\sqrt{\eta_\star \sin I}}, \\
        P_{II \to III} &= \frac{-2\pi \eta_\star \cos I + 4\sqrt{\eta_\star \sin
            I}} {8\sqrt{\eta_\star \sin I}}.
    \end{split}
\end{equation}

\subsection{Nontrivial Test Problem, ``Problem 1.5'' (sims)}\label{ss:p1_toy}

It is clear that $\eta'$ in \autoref{eq:p1_delta_eq} being the same sign along
both $\mathcal{C}_{\pm}$ is the source of guaranteed capture during
spindown. In order to test that \autoref{eq:p_c_gen} is not only qualitatively
correct but also quantitatively exact, we seek to simulate a problem with
probabilistic behavior. \textbf{Note: This section is probably superceded by
\autoref{ss:p1_spinup} or \autoref{s:p3}.}. We investigate test problem where
$\eta$ is varied following
\begin{equation}
    \eta' = \epsilon \eta \mu.
\end{equation}
This is a pure mathematical exercise, but resembles weak tidal friction in the
subsynchronous rotation regime.

Straightforward evaluation of \autoref{eq:p1_delta_eq} yields
\begin{equation}
    \Delta_{\pm} = \pm \epsilon \eta_\star \s{
        2\pi \sin I \pm 12 \sqrt{\eta_\star \sin I}\cos I
            + 2\pi \eta_\star \cos^2 I}.
\end{equation}
It bears noting that we must evaluate these integrals at $\eta = \eta_\star$ the
value at separatrix crossing. Since $\eta$ evolves over time, $\eta_{\star}$
depends on initial conditions. While one might expect the adiabatic invariance
of the action variable determines $\eta_\star$, this turns out to be slightly
inaccurate for reasons (I haven't worked this out convincingly yet).
% not be the case; we address this in \autoref{ss:noncons}.

Since $\abs{\Delta_+} > \abs{\Delta_-}$, separatrix encounters originating in
zone III always result in capture. However, encounters originating in zone I
experience capture probability
\begin{equation}
    P_c(\eta_\star) = \frac{24\cos I \sqrt{\eta_\star \sin I}}{
        2\pi \sin I + 12\sqrt{\eta_\star \sin I}\cos I
            + 2\pi \eta_\star \cos^2 I}.\label{eq:p_c_p1}
\end{equation}

To test this capture probability, we test capture probability at a variety of
$I$ for similar ICs. For fixed initial condition $(\mu_0, \phi_0)$ in zone I, we
vary the initial $\eta \in [0.01, 0.02]$ and $\epsilon \in 10^{[-3, -2]}$. In
this way, every simulation experiences probabilistic separatrix crossing at
roughly comparable times. We then repeated this procedure for three different
values $I = 10^\circ, 20^\circ, 25^\circ$. A natural spread in $\eta_\star$ was
observed at each $I$ value, which permitted comparison to \autoref{eq:p_c_p1}
independently. The results for the three $I$ values are presented in
\autoref{fig:2_1hist}, where good agreement with
\autoref{eq:p_c_p1} is shown.
\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{../initial/2_toy2/1hist10.png}
    \end{subfigure}

    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{../initial/2_toy2/1hist25.png}
    \end{subfigure}
    \caption{Simulations depicting $P_c(\eta_\star)$ at three different values
    $I = 10^\circ, 20^\circ, 25^\circ$ compared to analytical estimate
    \autoref{eq:p_c_p1}. The top panel is a stacked histogram showing the
    capture (into separatrix) and escape occurence counts at each $\eta_\star$.
    The bottom panel are the inferred estimates of $P_c(\eta_{\star})$ (capture
    probability) in blue and analytical curve in red. Uncertainties in
    $P_c(\eta_\star)$ from data are estimated as $\sqrt{N}$ where $N$ is the
    number of captures per bin. Also reported is the average value of
    $\eta_\star$.}\label{fig:2_1hist}
\end{figure}

% \subsection{Conservation of Adiabatic Invariant}\label{ss:noncons}

% Under the theory of adiabatic invariants (see e.g.\ \cite{henrard1982,
% l_and_l}), trajectories following Hamiltonian equations of motion save for small
% $\eta'$ conserve action variable $I$ except at separatrix crossing, where
% \begin{equation}
%     I \equiv \frac{1}{2\pi}\oint p\;\mathrm{d}q.
% \end{equation}
% $I$ has physical interpretation the enclosed phase space area of trajectories.
% Its evolution can be seen in \autoref{fig:2_1testo4}. Under the assumption $I$
% is conserved except at separatrix crossing, $\eta_\star$ is fixed implicitly by
% $I = A_{sep}(\eta_\star)$ for $I$ the initial value of the action variable.
% Since $A_{sep}$ is given by \autoref{eq:a_sep}, it would appear $\eta_\star$ can
% be predicted exactly.

% \textbf{NOTE:} The rest of this section is not very complete.

% This turns out to be accurate in \autoref{ss:p1_spindown} but not in
% \autoref{ss:p1_toy}, so it's probably non-physical/has to do with $\eta' = 0$
% when $\mu = 0$. Not sure whether this is physically useful. Useful formulas for
% $\rd{I}{t}$ can be taken from~\cite{henrard1982, l_and_l}:
% \begin{equation}
%     \rd{I}{\tau} = \s{\pd{H}{\eta}T(\eta) - \int\limits_0^{T(\eta)}
%             \pd{H}{\eta}\;\mathrm{d}\tau} \eta'
%         = -\pd{}{w}\p{\pd{S}{\eta}}\eta'.
% \end{equation}
% Here, $T(\eta)$ is the period of the orbit, $w$ is the angle variable conjugate
% to $I$, and $S$ is the generating function for canonical transformation $\mu =
% \pd{S}{\phi}, w = \pd{S}{I}$. These expressions are hard to evaluate, but maybe
% they will tell us how much to expect $\rd{I}{\tau}$ to deviate from zero near
% the separatrix?

\section{Problem 2: Weakly Dissipitive Hamiltonian}\label{s:p2}

\subsection{$\eta < \eta_c$, Separatrix Encounters (sims)}\label{ss:p2_lt}

We next consider a toy separatrix crossing problem in the $\dot{\eta} = 0$
limit, the case discussed in \autoref{ss:meln}. Consider adding to the
Hamiltonian Cassini State problem an additional aligning term
$\p{\rd{\theta}{t}}^{(1)} = -\frac{1}{t_s}\sin\theta$ that favors alignment
$\theta = 0$ on synchronization timescale $t_s$. This translates to equation of
motion
\begin{equation}
    \rd{\hat{s}}{\tau} = \p{\hat{s} \cdot \hat{l}}\p{\hat{s} \times \hat{l}}
        - \eta\p{\hat{s} \times \hat{l}_p}
        - \epsilon \sin \theta \hat{\theta}.
\end{equation}
We have divided \autoref{eq:dsdt_base} by $\alpha$ and defined $\tau \equiv
\alpha t, \epsilon \equiv \frac{1}{\alpha t_s}$. The equations of motion for
canonical variables $\mu \equiv \cos\theta, \phi$ are
\begin{subequations}\label{se:p2_eom}
    \begin{align}
        \phi' &= -\cos\theta
            + \eta\p{\cos I + \sin I \frac{\mu}{\sqrt{1 - \mu^2}} \cos \phi},\\
        \mu' &= -\eta \sin I \sin \theta \sin \phi
            \s{+ \epsilon\p{1 - \mu^2}}.
    \end{align}
\end{subequations}
Primes denote $\rd{}{\tau}$. The bracketed term in $\mu'$ is the perturbation
component. We will take $\alpha, \epsilon > 0$, which will result in separatrix
encounters originating in zone III, but an identical procedure for $\epsilon <
0$ describes separatrix encounters originating in zone I.

With a dissipitive term that favors alignment, we expect that separatrix
encounters can only occur from zone II and zone III\@. We will focus on the
outcomes of trajectories origiating in zone III for now. We now aim to compute
\autoref{eq:p_c_gen}, for which we need:
\begin{align}
    \Delta_\pm &= \int\limits_{\mathcal{C}_\pm}
            \phi'\epsilon\p{1 - \mu^2}\;\mathrm{d}t,\nonumber\\
        &= \mp 2\pi\p{1 - 2\eta \sin I - \eta^2 \cos^2 I}
                + 16\cos I \eta^{3/2}\sqrt{\sin I}.\label{eq:p2_delta}
\end{align}
Note that along $\mathcal{C}_{\pm}$ that $\mathrm{d}\phi$ has sign $\mp$
respectively. From these two integrals, we obtain
\begin{equation}
    P_c = \frac{16\eta^{3/2}\cos I \sqrt{\sin I}}{\pi
        \p{1 - 2\eta \sin I - \eta^2 \cos^2 I}
        + 8\eta^{3/2}\cos I \sqrt{\sin I}}.\label{eq:p1pc}
\end{equation}

@@

A further interesting feature in \autoref{fig:0_3stats} is a small portion of
points in zone II near CS4  at $\phi = 2\pi$ that escape. These points accrue
$\Delta_+$ on their first orbit and escape without librating. Visually, this
zone is obvious in \autoref{fig:0_6manifolds} as the separation between
$W_s^{(1)} \cup W_u^{(1)}$ and $W_s^{(0)} \cup W_u^{(0)}$ near $\phi = 2\pi$.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_1.png}
    \caption{Plot of the CSs that initial conditions converge to. Each subplot
    has a set of initial conditions (black dots) and the CS (red dot) that each
    of these initial conditions converge to. Initial conditions are chosen
    uniformly in $(\mu, \phi)$ coordinates. Key features to note are secure
    convergence to CS1 for all zone I initial conditions, probabilistic
    convergence to CS1 or CS2 for zone III initial conditions (first half of
    \autoref{ss:p2_lt}), and secure convergence to CS2 for all zone II initial
    conditions (second half of \autoref{ss:p2_lt}). The parameters for the
    simulation are in the title.}\label{fig:0_3stats}
\end{figure}
